## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10 

if  (a<=5){
 print(a)
} else {
    print("Its more than 5")
  }
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 15

if  (a<=5){
 print("Small")
} else if (a > 5 & a <= 10) {
    print("Medium")
} else if (a > 10 | a == 7){
    print("Large")
}else {
    print ("Huge")
  }
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <- 5

if (a/2 != 0) {
  if (a == 5){
    print ("a is 5")
  }
}
    
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5){
  print(paste(value, "is greater than", value - 1))
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <-c(1,1,1,1,1,1,1)

for (i in c(1,3,6)){
  a_vector [i] <- 6
}

a_vector
```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow = 2, ncol = 2)

for (row in 1:2){
  for (col in 1:2){
    mat[row, col] <-0
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b<- 0
while (b < 10){
  b <- b + 2
  print (b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}
```

An example:

```{r}
test_temp <- c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735)

c_to_f <- function (dat){
  temp_fdhr <- dat*(9/5) + 32
  
  return (temp_fdhr)
}

print (c_to_f(26))
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_fdhr <- c_to_f(test_temp)
data_fdhr
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow <- 2)

apply(test_mat, 2, sum)
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735), nrow = 2)

apply(test_temp_mat, 1, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    
  #Codewars ID: Adharia
 # 1) Write a function named setAlarm/set_alarm/set-alarm/setalarm (depending on language) which receives two parameters. The first parameter, employed, is true whenever you are employed and the second parameter, vacation is true whenever you are on vacation.

#The function should return true if you are employed and not on vacation (because these are the circumstances under which you need to set an alarm). It should return false otherwise. Examples:
    
  set_alarm <- function(employed, vacation) {
  if (employed && !vacation) {
    return(TRUE)
  } else {
    return(FALSE)
  }
  }
 # 2) Write a function which converts the input string to uppercase.
    
    makeUpperCase <- function(s) {
  upperCaseString <- toupper(s)
  return(upperCaseString)
}

# 3) Write a function which calculates the average of the numbers in a given list.

#Note: Empty arrays should return 0.
    
    find_average <- function(vec){
  if (length(vec) == 0) {
    return(0)
  }
  
  average <- sum(vec) / length(vec)
  
  return(average)
    }
    
# 4) You will be given an array a and a value x. All you need to do is check whether the provided array contains the value.

#Array can contain numbers or strings. X can be either.

#Return true if the array contains the value, false if not.
    
  check <- function(a, x) {
  result <- x %in% a
  return(result)
  }
  
# 5) In this simple assignment you are given a number and have to make it negative. But maybe the number is already negative?
  
  make_negative <- function(num){
  if(num >= 0){
    return(-num)
  } else {
    return(num)
  }
}

    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
fizzbuzz <- function(number) {
  if (number %% 3 == 0 && number %% 5 == 0) {
    return("fizzbuzz")
  } else if (number %% 3 == 0) {
    return("fizz")
  } else if (number %% 5 == 0) {
    return("buzz")
  } else {
    return(as.character(number))
  }
}

print(fizzbuzz(3)) 
print(fizzbuzz(15))
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.
    
    marioStairs <- function(length) {
  for (i in 1:length) {
    cat(rep("X", i), "\n")
  }
    }
    
    marioStairs(4)
    ```
